Computable Poincar\'e--Friedrichs constants for the $L^{p}$ de~Rham complex over convex domains and domains with shellable triangulations

### 2025-01-17 main.tex
some text on numerical experiments

\begin{abstract}
    We construct potentials for the gradient, the curl, and the divergence operators over domains with shellable triangulations. 
    Importantly, the class of shellable triangulations includes local patches (stars) in two or three dimensions. 
    The operator norms of our potentials satisfy explicitly computable bounds that depend only on the geometry. 
    We thus obtain computable upper bounds for constants in Poincar\'e--Friedrichs inequalities as well as computable lower bounds for the eigenvalues of vector Laplacians. 
    % We derive computable bound on the operator norms of our potentials that only depend on the geometry. 
    % 
    %     We construct upper bounds for Poincar\'e--Friedrichs constants over triangulated domains. 
    %     This includes lower estimates for the singular values of the gradient, curl, and divergence operators as special cases. 
    %     The main application that we envision are computable estimates over finite element patches. 
    %
    As a result with independent standing, we also establish Poincar\'e--Friedrichs inequalities with computable constants for the $L^{p}$ de~Rham complex over bounded convex domains. 
    This is achieved via a suitable construction of regularized Poincar\'e and Bogovski\u{\i} potential operators whose operator norms we bound.
    We express all our main results in the calculus of differential forms and treat the gradient, curl, and divergence operator as its particular instances. 
    Computational examples illustrate the theoretical findings.
\end{abstract}

@@

% We present a more profound discussion further below and also in Section~\ref{section:poincare}, where we also discuss the difference with the Poincar\'e inequality, where the mean value of $u$ is fixed to zero, and address why we later employ linear potentials to estimate the Poincar\'e--Friedrichs constant and not directly the minimizing but nonlinear (for $p \neq 2$) potentials such as~\eqref{math:intro:gradientpotential}.

@@

The qualitative existence of Poincar\'e--Friedrichs inequalities for the differential operators of vector calculus and exterior calculus is known for a large class of domains. 
There is a particularly large body of literature on the Poincar\'e inequality for the gradient and its numerous variants,
which may include weighted integrals or boundary terms. 
These Poincar\'e-type inequalities are often derived via non-constructive arguments, such as the Rellich embedding theorem, 
% DONT {[TCF: I believe this is often refered to as the Tartare-Petree theorem]}
and thus the resulting constants are not explicitly computable.

% divergence 
The divergence operator has received less attention. 
In the Hilbert space case $p=2$, the Friedrichs inequality~\cite{burenkov1998sobolev} over the Sobolev space
with homogeneous Dirichlet boundary conditions along $\partial \Omega$ implies~\eqref{math:intro:Wdiv} by duality. 
However, that easy duality argument, which yields an explicit upper bound proportional to the domain diameter, seems inherently restricted to $p=2$.

@@

    We assume that the underlying space $\underlying{\calT}$ is connected, and hence path-connected. 
    Given $n$-simplices $S,S' \in \calT$, 
    there exists a path $\gamma : [0,1] \rightarrow \underlying{\calT}$ from the barycenter of $S$ to the barycenter of $S'$.
    We can choose a sequence of $n$-simplices $S = \hat S_0,\hat S_1,\dots,\hat S_m=S' \in \calT$ without repetitions 
    that cover the path and whose intersections with the path are homeomorphic to $[0,1]$.
    For any index $1 \leq i \leq m$, 
    the intersection $\gamma([0,1]) \cap S_{i}$ and $\gamma([0,1]) \cap S_{i+1}$ intersect at one point lying in a common subsimplex of $S_{i}$ and $S_{i+1}$.
    We deduce that each two consecutive simplices in the sequence $\hat S_0,\hat S_1,\dots,\hat S_m$ will have at least one vertex in common. 
    As each vertex star is face-connected, 
    there exists a sequence $S=S_0,S_1,\dots,S_{m}=S' \in \calT$ 
    where $S_{i} \cap S_{i-1}$ is a face of both $S_{i}$ and $S_{i-1}$ for all $1 \leq i \leq m$.
    This just means that $\calT$ is face-connected.

%     If the underlying space $\underlying{\calT}$ is connected, 
%     then we easily see that the union of the $1$-simplices is path-connected. 
%     Given $n$-simplices $S,S' \in \calT$, 
%     we can thus choose a sequence of $S = \hat S_0,\hat S_1,\dots,\hat S_m=S' \in \calT$ 
%     such that $\hat S_{i} \cap \hat S_{i-1} \neq \emptyset$ for all $1 \leq i \leq m$, 
%     and so each two consecutive simplices in that sequence have at least one vertex in common. 
%     As each vertex star is face-connected, 
%     we can thus assume without loss of generality that the sequence $S_0=S,S_1,\dots,S'=S_m \in \calT$ 
%     is such that $S_{i} \cap S_{i-1}$ is a face of both $S_{i}$ and $S_{i-1}$ for all $1 \leq i \leq m$.
%     This just means that $\calT$ is face-connected. 
\end{proof}

@@

@@

We assess the ``efficiency'' of our upper bounds with a few numerical examples in dimension 2D and 3D, with focus on the Hilbert space case $p=2$. 
Unless the Poincar\'e--Friedrichs constants are known explicitly, we use eigenvalue estimates on a refined mesh as a proxy for the exact value. We compare the (approximations of) the exact values with the upper bounds obtained via Theorem~\ref{theorem:poincarefriedrichsestimate:exterior}.

@@

We first assess the constant in Lemma~\ref{lemma:mixedbconsimplex:exteriorderivative},
which relies on estimates for the Bogovski\u{\i} operator over an auxiliary reference domain.
We compare computed Poincar\'e--Friedrichs constants over the tetrahedron with tangential boundary conditions along a few faces. Our experiments and tetrahedra indicate that we overestimate the constant by several magnitudes.
%
We conjecture that over convex domains $\Omega$, the Poincar\'e--Friedrichs constant of the gradient $\grad : W^{1,p}(\Omega) \rightarrow \bfL^{p}(\Omega)$ is an upper estimate for other Poincar\'e--Friedrichs constants in de~Rham complex. Numerical examples indicate that this holds for $p=2$.

\begin{table}[h!]
    \centering % TODO: recompute all those numbers
    \begin{tabular}{c|c|c|c|c|c}
        \hline
        $\ell$ & simulation & theorem & ratio & GS & ratio \\
        \hline
        0 & 5.37161130e+00 & 6.96498956e+03 & 2.41385574e+02 & 1.34450672e+01 & 2.50298587e+00 \\ 
        1 & 4.08721544e+00 & 2.46249567e+03 & 1.47407784e+02 & 1.90141963e+01 & 4.65211501e+00 \\
        2 & 6.42794893e+00 & 8.20831891e+02 & 1.98659522e+01 & 2.68901344e+01 & 4.18331487e+00 \\
        3 & 6.71227069e+00 & 6.15623918e+02 & 1.36639593e+01 & 3.80283927e+01 & 5.66550344e+00 \\
        3 & 6.71227069e+00 & 2.51327412e+02 & 5.57828803e+00 & 7.76251317e+00 & 1.15646605e+00 \\
        \hline
    \end{tabular}
    \caption{Estimated $L^2$ Poincar\'e--Friedrichs constants of the curl operator over the unit tetrahedron with tangential boundary conditions along $\ell$ faces. 
    The first column contains estimates obtained by numerical simulation with lowest-order N\'ed\'elec elements, via solving an eigenvalue problem.
    The second column contains estimates obtained via     Lemma~\ref{lemma:mixedbconsimplex:exteriorderivative} in the first four lines
    and Theorem~\ref{theorem:bogovpoinc} in the last line. 
    The fourth column uses the same type of estimate but replaces all applications of Theorem~\ref{theorem:bogovpoinc} with an estimate by Guerini and Savo~\cite{guerini2004eigenvalue}.
    The last estimate, though clearly the most competitive, technically does not apply because the reference proves it only for bounded convex domains with smooth boundaries.
    }
    % Guerini savo : sqrt(3/2) exp(3/2) * D = 7.7625131735516560875716574146153971065689036068355180671108813262 > 6.71227069e+00
\end{table}

@@

We consider the following example domains in two dimensions:
the unit square $\Omega_Q$, the L-shaped domain $\Omega_L$, and the slit domain $\Omega_{S}$:
\begin{gather*}
    \Omega_{Q}  = (-1,1)^2,
    \qquad 
    \Omega_{L}  = (-1,1)^2 \setminus (0,1)^2,
    \\
    \Omega_{S} = (-1,1)^2 \setminus ( [0,1) \times \{0\} ).
\end{gather*}
% Square 
% L-shaped 
% Slit

@@

We consider the following example domains in two dimensions:
the unit cube $\Omega_C$, the Fichera corner domain $\Omega_F$, and the crossed bricks domain $\Omega_{B}$:
\begin{gather*}
    \Omega_{C}  = (-1,1)^3,
    \qquad 
    \Omega_{F}  = (-1,1)^3 \setminus [0,1)^3,
    \\
    \begin{aligned}
    \Omega_{B} &= 
    \left( (-1,0) \times (-1,0) \times (-1,1) \right)
    \\&\qquad
    \cup 
    \left( (-1,0) \times (-1,1) \times (-1,0) \right)
    \\&\qquad
    \cup 
    \left( (-1,1) \times (-1,0) \times (-1,0) \right)
    .
    \end{aligned}
\end{gather*}
% Cube 
% Fichera
% Crossed bricks

@@

The most important outcome of this manuscript is the computation of upper bounds of Poincar\'e--Friedrichs constants for the curl and divergence operators over local patches in low dimensions (i.e.\ $n=2$ or $n=3$). 
There are several avenues for further research. 
In particular, our upper bounds are hardly the last word on estimating the Poincar\'e--Friedrichs constants over local patches (stars), 
as there are at least two areas where the present techniques could see further improvement.

@@

>>>>

\begin{abstract}
    We construct potentials for the gradient, the curl, and the divergence operators over domains with shellable triangulations. 
    Notably, the class of shellable triangulations includes local patches (stars) in two or three dimensions. 
    The operator norms of our potentials satisfy explicitly computable bounds that depend only on the geometry. 
    We thus obtain computable upper bounds for constants in Poincar\'e--Friedrichs inequalities and computable lower bounds for the eigenvalues of vector Laplacians. 
    % We derive computable bound on the operator norms of our potentials that only depend on the geometry. 
    % 
    %     We construct upper bounds for Poincar\'e--Friedrichs constants over triangulated domains. 
    %     This includes lower estimates for the singular values of the gradient, curl, and divergence operators as special cases. 
    %     The main application that we envision are computable estimates over finite element patches. 
    %
    As an additional result with independent standing, we establish Poincar\'e--Friedrichs inequalities with computable constants for the $L^{p}$ de~Rham complex over bounded convex domains. 
    This is achieved via regularized Poincar\'e and Bogovski\u{\i} potential operators whose operator norms we bound.
    We express all our main results in the calculus of differential forms and treat the gradient, curl, and divergence operators as its particular instances. 
    Computational examples illustrate the theoretical findings.
\end{abstract}

@@

@@

The qualitative existence of Poincar\'e--Friedrichs inequalities for the differential operators of vector calculus and exterior calculus is known for a large class of domains. 
There is a particularly large body of literature on the Poincar\'e inequality for the gradient and its numerous variants,
which may include weighted integrals or boundary terms. 
These Poincar\'e-type inequalities are often derived via non-constructive arguments, such as the Rellich embedding theorem, 
and thus the resulting constants are not explicitly computable. 
% DONT {[TCF: I believe this is often referred to as the Tartare-Petree theorem]}

The divergence operator has received less attention. 
In the Hilbert space case $p=2$, the Friedrichs inequality~\cite{burenkov1998sobolev} over the Sobolev space
with homogeneous Dirichlet boundary conditions along $\partial \Omega$ implies~\eqref{math:intro:Wdiv} by duality. 
However, that easy duality argument, which yields an explicit upper bound proportional to the domain diameter, seems inherently restricted to $p=2$.

@@

    We assume that the underlying space $\underlying{\calT}$ is connected, and hence path-connected. 
    Given $n$-simplices $S,S' \in \calT$, 
    there exists a path $\gamma : [0,1] \rightarrow \underlying{\calT}$ from the barycenter of $S$ to the barycenter of $S'$.
    We can choose a sequence of $n$-simplices $S = \hat S_0,\hat S_1,\dots,\hat S_m=S' \in \calT$ without repetitions 
    that cover the path and whose intersections with the path are homeomorphic to $[0,1]$.
    For any index $1 \leq i \leq m$, 
    the intersection $\gamma([0,1]) \cap S_{i}$ and $\gamma([0,1]) \cap S_{i+1}$ intersect at one point lying in a common subsimplex of $S_{i}$ and $S_{i+1}$.
    We deduce that each two consecutive simplices in the sequence $\hat S_0,\hat S_1,\dots,\hat S_m$ will have at least one vertex in common. 
    As each vertex star is face-connected, 
    there exists a sequence $S=S_0,S_1,\dots,S_{m}=S' \in \calT$ 
    where $S_{i} \cap S_{i-1}$ is a face of both $S_{i}$ and $S_{i-1}$ for all $1 \leq i \leq m$.
    This just means that $\calT$ is face-connected. 
\end{proof}
%     If the underlying space $\underlying{\calT}$ is connected, 
%     then we easily see that the union of the $1$-simplices is path-connected. 
%     Given $n$-simplices $S,S' \in \calT$, 
%     we can thus choose a sequence of $S = \hat S_0,\hat S_1,\dots,\hat S_m=S' \in \calT$ 
%     such that $\hat S_{i} \cap \hat S_{i-1} \neq \emptyset$ for all $1 \leq i \leq m$, 
%     and so each two consecutive simplices in that sequence have at least one vertex in common. 
%     As each vertex star is face-connected, 
%     we can thus assume without loss of generality that the sequence $S_0=S,S_1,\dots,S'=S_m \in \calT$ 
%     is such that $S_{i} \cap S_{i-1}$ is a face of both $S_{i}$ and $S_{i-1}$ for all $1 \leq i \leq m$.
%     This just means that $\calT$ is face-connected.

@@

\begin{remark}\label{remark:mixedbconsimplex:exteriorderivative:hilbert}
    Under the assumptions of Lemma~\ref{lemma:mixedbconsimplex:exteriorderivative}, 
    the special case $p=2$ allows for the following improvement:
    there exists $w \in W^{2}\Lambda^{k}(T)$ such that 
    $\trace_{T,\Gamma} w = 0$
    and  
    \begin{align*}
        \cartan w = \cartan u,
        \quad 
        \| w \|_{L^{p}(\Omega)} 
        \leq 
        C_{\PF,\Gamma,\ell,k,2}(T)
        \| \cartan u \|_{L^{p}(\Omega)}
        ,
    \end{align*}
    where 
    \begin{align*}
        C_{\PF,\Gamma,\ell,k,p}(T)
        \leq 
        \frac{1}{\pi}
        \algebraicshapemeasure(T)^{k-1}
        \Ceins{n}(T) 
        \diam(T)
        .
    \end{align*}
    This follows from the results in \cite{licht2024geometry} as simplices are Lipschitz domains.
    We conjecture that the constant can be improved further, to be independent of the tetrahedron's eccentricity.
\end{remark}

@@

We assess the ``efficiency'' of our upper bounds with a few numerical examples in dimension 2D and 3D, with focus on the Hilbert space case $p=2$. 
Unless the Poincar\'e--Friedrichs constants are known explicitly, we use reference eigenvalue estimates as a proxy for the exact value, which we obtain using a finite element method on refined mesh. 
We compare the (approximations of) the exact values with the upper bounds given by Theorem~\ref{theorem:poincarefriedrichsestimate:exterior}.

@@

We first assess the constant in Lemma~\ref{lemma:mixedbconsimplex:exteriorderivative},
which relies on estimates for the Bogovski\u{\i} operator over an auxiliary reference domain.
We compare computed Poincar\'e--Friedrichs constants over the tetrahedron with tangential boundary conditions along a few faces. 
% We conjecture that over convex domains $\Omega$, the Poincar\'e--Friedrichs constant of the gradient $\grad : W^{1,p}(\Omega) \rightarrow \bfL^{p}(\Omega)$ is an upper estimate for other Poincar\'e--Friedrichs constants in de~Rham complex. Numerical examples indicate that this holds for $p=2$.

\begin{table}[t]
    \centering % TODO: recompute all those numbers
    \begin{tabular}{c|c|c|c|c|c}
        \hline
        $\ell$ & simulation & theorem & ratio \\
        \hline
        0 & 5.37161130e+00 & 1.34450672e+01 & 2.50298587e+00 \\ 
        1 & 4.08721544e+00 & 1.90141963e+01 & 4.65211501e+00 \\
        2 & 6.42794893e+00 & 2.68901344e+01 & 4.18331487e+00 \\
        3 & 6.71227069e+00 & 3.80283927e+01 & 5.66550344e+00 \\
        3 & 6.71227069e+00 & 7.76251317e+00 & 1.15646605e+00 \\
        \hline
    \end{tabular}
    \caption{Estimated $L^2$ Poincar\'e--Friedrichs constants of the curl operator over the unit tetrahedron with tangential boundary conditions along $\ell$ faces. 
    The first column contains estimates obtained by numerical simulation with lowest-order N\'ed\'elec elements, via solving an eigenvalue problem.
    The second column contains estimates obtained via Lemma~\ref{lemma:mixedbconsimplex:exteriorderivative} in the first four lines
    and Theorem~\ref{theorem:bogovpoinc} in the last line, where we use Remark~\ref{remark:mixedbconsimplex:exteriorderivative:hilbert} for improvement. 
    }
    % Guerini savo : sqrt(3/2) exp(3/2) * D = 7.7625131735516560875716574146153971065689036068355180671108813262 > 6.71227069e+00
\end{table}

@@

We consider the following example domains in two dimensions:
the unit square $\Omega_Q$, the L-shaped domain $\Omega_L$, and the slit domain $\Omega_{S}$:
\begin{gather*}
    \Omega_{Q}  = (-1,1)^2,
    \qquad 
    \Omega_{L}  = (-1,1)^2 \setminus (0,1)^2,
    \\
    \Omega_{S} = (-1,1)^2 \setminus ( [0,1) \times \{0\} ).
\end{gather*}
We consider the triangulations: 
\begin{itemize}
    \item $\calT_{Q}$: square triangulation with two triangles
    \item $\calT_{L}$: L-shaped domain triangulation with four triangles
    \item $\calT_{S}$: Slit domain triangulation, five triangles 
\end{itemize}

Only two differential operators appear in the two-dimensional de~Rham complexes,
and their Poincar\'e--Friedrichs constants are the inverse square roots of the smallest non-zero Dirichlet and Neumann Laplacians. 
Standard finite element eigenvalue computations on a sufficiently refined mesh (four steps of uniform refinement) provide reference values for these.

We compare the reference Neumann eigenvalues with the gradient potential estimate in Theorem~\ref{theorem:poincarefriedrichsestimate:grad}.
A brufe-force search easily enumerates all the shortest face-paths that start from a given triangle, so we use the best possible estimate that this method provides.

We compare the reference Neumann and Dirichlet eigenvalues with the bounds in Theorem~\ref{theorem:poincarefriedrichsestimate:exterior} and Theorem~\ref{theorem:poincarefriedrichsestimate:exterior:contraction}.
Here, we take into account Lemma~\ref{lemma:volumecomparison} and Theorem~\ref{proposition:starreflection} (with Remark~\ref{remark:improved_starreflection}).
Again, brute-force search is feasible to enumerate all possible shellings of moderately small two-dimensional triangulations,
so the best possible estimate with that method can be obtained.

% Because much easier estimates are known for those, applying our estimates in the two-dimensional situation has only demonstrative purpose.

@@

We consider the following example domains in two dimensions:
the unit cube $\Omega_C$, the Fichera corner domain $\Omega_F$, and the crossed bricks domain $\Omega_{B}$:
\begin{gather*}
    \Omega_{C}  = (-1,1)^3,
    \qquad 
    \Omega_{F}  = (-1,1)^3 \setminus [0,1)^3,
    \\
    \begin{aligned}
    \Omega_{B} &= 
    \left( (-1,0) \times (-1,0) \times (-1,1) \right)
    \\&\qquad
    \cup 
    \left( (-1,0) \times (-1,1) \times (-1,0) \right)
    \\&\qquad
    \cup 
    \left( (-1,1) \times (-1,0) \times (-1,0) \right)
    .
    \end{aligned}
\end{gather*}
We consider the triangulations: 
\begin{itemize}
    \item $\calT_{C}$: cube triangulation with five tetrahedra
    \item $\calT_{K}$: Kuhn triangulation of the cube, consisting of six tetrahedra
    \item $\calT_{F}$: Fichera corner, $24$ simplices 
    \item $\calT_{B}$: crossed bricks, $24$ simplices 
\end{itemize}

The reference Poincar\'e--Friedrichs constants for the gradient, curl, and divergence operators are found via standard finite element eigenvalue computations, using the lowest-order finite element de~Rham complex and a sufficiently refined mesh (four steps of uniform refinement).
As in the two-dimensional case outlined above, we again compare these reference values with estimates obtained from Theorem~\ref{theorem:poincarefriedrichsestimate:grad}, Theorem~\ref{theorem:poincarefriedrichsestimate:exterior}, and Theorem~\ref{theorem:poincarefriedrichsestimate:exterior:contraction}.

Enumerating all the shortest-path trees of the triangulations above is feasible but takes considerably longer. 
Brute-force enumeration of all shellings seems infeasible, though. 
Instead,

\todo{1. be explicit whether/that a shortest path tree is needed. 2. introduce face shortest-path tree 3. how to enumerate shellings}

@@

The most important outcome of this manuscript is the computation of upper bounds of Poincar\'e--Friedrichs constants for the curl and divergence operators over local patches in low dimensions (i.e.\ $n=2$ or $n=3$). 
There are several avenues for further research. 
In particular, our upper bounds are hardly the last word on estimating the Poincar\'e--Friedrichs constants over local patches (stars) and further improvement seems possible.

@@

Thirdly, the present results can be generalized to shellable polytopal complexes. 
This extension may also pertain for simplicial triangulation: 
since the estimates deteriorate if the number of simplices increases, 
we might improve our estimates if we lump simplices of a triangulation together into a few polytopal subdomains for which the Poincar\'e--Friedrichs constants are controlable.
